Testing and Isolation Efficacy:\\Insights from a Simple Epidemic Model

### 2022-03-07 manuscript_BMB.tex
added Muler to refs, marked %reviewer:com in the mans

\newcommand{\fref}[1]{Fig.~\ref{#1}}
\newcommand{\appref}[1]{Appendix~\ref{app:#1}}
% \newcommand{\Rlogo}{R\xspace}
\newcommand{\percap}{\emph{per capita}\xspace}
\newcommand{\Rnum}{\ensuremath{\mathcal{R}_0}\xspace}
\newcommand{\covid}{COVID-19\xspace}
\newcommand{\pro}[1]{\ensuremath{\frac{\partial #1}{\partial \rho}}}
\newcommand{\pder}[2]{\ensuremath{\frac{\partial#1}{\partial#2}}} %\pder y{x}
\newcommand{\testing}[1]{\ensuremath{{\cal T}_{#1}}\xspace}
\newcommand{\testinghat}[1]{\ensuremath{\hat{\cal T}_{#1}}\xspace}
\newcommand{\testtarget}{\ensuremath{w\_{IS}}}
\newcommand{\oldtesttarget}{\ensuremath{\frac{w_I}{w_S}}\xspace}
\newcommand*\subtxt[1]{_{\textnormal{#1}}}
\newcommand{\rev}[1]{\comment{red}{Revised}{#1}}
\renewcommand{\rev}[1]{{#1}}
%% Roman subscript
\DeclareRobustCommand\_{\ifmmode\expandafter\subtxt\else\textunderscore\fi}

@@

Testing individuals for pathogens can affect the spread of epidemics. Understanding how individual-level processes of sampling and reporting test results can affect community- or population-level spread is a dynamical modeling question.
The effect of testing processes on epidemic dynamics depends on factors underlying implementation, particularly testing intensity and on whom testing is focused.
Here, we use a simple model to explore how the individual-level effects of testing might directly impact population-level spread.
Our model development was motivated by the \covid epidemic, but has generic epidemiological and testing structures. 
To the classic SIR framework we have added a \percap testing intensity, and compartment-specific testing weights, which can be adjusted to reflect different testing emphases --- surveillance, diagnosis, or control. 
We derive an analytic expression for the relative reduction in the basic reproductive number due to testing, test-reporting and related isolation behaviours.
Intensive testing and fast test reporting are expected to be beneficial at the community level because they can provide a rapid assessment of the situation, identify hot spots, and may enable rapid contact-tracing. 
Direct effects of fast testing at the individual level are less clear, and may depend on how individuals' behaviour is affected by testing information.
Our simple model shows that under some circumstances both increased testing intensity and faster test reporting can \emph{reduce} the effectiveness of control, and allows us to explore the conditions under which this occurs.
Conversely, we find that focusing testing on infected individuals always acts to increase effectiveness of control.

@@

The Disease-Free Equilibrium (DFE) for the expanded SIR model (Eqs.~\ref{model}) is found by setting the infected compartments to 0 and solving for the unknowns. 
% reviewer1:com3 
\rev{The DFE depends on \percap testing intensity $\rho$, the rate of test return $\omega$, and the population size $N$.} 
The DFE is 
\begin{equation}
\label{dfe}
S\_n^*= \frac{\rho}{\omega} N, \quad S\_u^*= N-S\_n^*, \quad\text{and}\quad I_j=R_j=0 \ \text{for all }j.
\end{equation}
The corresponding \percap testing rate (Eq.~\ref{test}) for the infected compartment $I$ at DFE is one of the key analysis parameters and can be simplified as 
\begin{equation}
\label{eq:fi}
\testinghat{I} = (\omega\rho/(\omega-\rho))w_I/w_S \quad .
\end{equation}

@@

In this paper, we have developed and analyzed a simple compartmental model that combines epidemiological dynamics---as defined by a simple SIR model---with the dynamics of testing and isolation. Our model is a caricature: it models the most basic feedbacks between epidemic and testing processes, but does not attempt to incorporate the many known complications of COVID-19 epidemiology (e.g., exposed, pre-symptomatic, and asymptomatic compartments \citep{kain2021chopping}; time-varying testing rates; behavioural dynamics \citep{weitz2020awareness}); 
% reviewer2:com1
\rev{Further, our model ignores the fact that the testing and isolation parameters can change over time due to public health responses. In other words, our model makes an assumption of ``independence'': we assume that testing and isolation affect the epidemic progression, but we do not account for the fact that epidemic progression is also likely to affect testing and isolation strategies.}
Thus, it is most appropriate for assessing the \emph{qualitative} phenomena that arise from the interactions between transmission dynamics and testing, rather than for making quantitative predictions or guiding pandemic responses.

@@

This finding does, however, suggest that cautious behaviour by people waiting for test results might have appreciable effects on epidemic spread. This reinforces the idea that people waiting for test results should be urged to isolate and take other measures to prevent infecting other, or being infected. It also provides general support for the idea of institutional ``gateway'' testing which restricts individuals' contacts or access until one or more negative tests have been received \cite{Muller}.

@@

Although we model the testing process in more detail than typical epidemiological models, one place where more detail could be informative is in the processes determining the testing weights $\{w_S, w_I, w_R\}$. While random testing, as done for surveillance purposes, unambiguously leads to equal testing weights, making precise quantitative connections between public-health practices and testing weights is difficult in other contexts. The testing weights reflect the correlation between an individual's risk of infection and their likelihood of being tested due to age, occupation, geographic location, etc.. This correlation is influenced, among many other factors, by the proportion of the uninfected population with COVID-like symptoms (e.g., due to seasonal upper respiratory tract infections); the concentration of transmission and testing in hot spots such as long-term care facilities and high-density workplaces; the overall testing intensity (and hence, e.g., restriction to symptomatic individuals); and the proportion of COVID-infected people who are symptomatic. Future steps should explore mathematically tractable ways to model some of these factors more precisely. For example, separating the infected class into exposed, symptomatic, and a- or pre-symptomatic compartments and allowing the testing weights to vary across non-symptomatic (exposed/asymptomatic/presymptomatic) vs.\ symptomatic compartments could reflect the allocation of tests for diagnostic purposes (targeting symptomatic individuals) vs.\ contact-tracing (targeting infected but non-symptomatic individuals) vs.\ screening (relatively equal weights, depending on the venue). Alternatively, one could make the testing weights depend on the testing intensity or test-return rate as suggested above. Whatever complexity is added would probably put the model beyond reach of the analytical methods we have used in this paper, but one could still use semi-numerical methods such as constructing the next-generation matrix and using it to evaluate the derivatives of $\Rnum$ with respect to the parameters numerically.

@@

%% \fred{ our model makes the counterintuitive suggestion that increased testing rate could increase $\Rnum$. However, note that in our model, we made the assumption that once an individual is in the ``waiting'' compartment, they cannot be tested again until the test result returns. Thus, when the test return time is very long, this creates a massive time-delay where an individual may become infected while waiting for the test result, and then the test result comes back as negative, and so they believe they are negative when they are actually positive (and so they do not self-isolate at all, under the belief that they are negative).
%% So, in this case, it would be beneficial to actually re-test people even if their first test result has not come back. In fact, one could probably conceive of ``testing breadth'' (which roughly corresponds to the number of different people you test over some period of time) vs. ``testing depth (which corresponds to the number of times you test a single person over a period of time). Note that testing breadth gives us granular information ``in space but not in time'' (i.e., it tells us how many people are or are not infected at a single point in time), whereas testing depth gives us granular information ``in time but not in space''. The pathological situation in our model can be viewed as an imbalance between testing depth and breadth, where the testing strategy is overly-broad but not sufficiently deep.
%% In particular, one could build models that differentiate/partition ``testing intensity'' into ``testing breadth'' and ``testing depth''. In such a model, there should always exist some allocation of breadth and depth that makes it so that increasing the overall testing intensity is beneficial (i.e., if we choose the right strategy, more testing should always be good).
%% The most important point of all this: while the exact effect we identified in our model (that more testing can increase $\Rnum$ if done incorrectly) is unlikely to happen in real life, a weaker ``sister effect'' probably happens in real life all the time: we may be allocating our testing resources in the wrong breadth vs. depth ratio. Our model ``proves'' that pure-breadth is not always the optimal strategy (because, in fact, it can increase $\Rnum$). So, a more realistic future question is: what is the optimal allocation between breadth and depth that decreases $\Rnum$ most sharply for a given overall testing intensity? For example, in our model, when we increase testing intensity, we are mostly increasing breadth. So, even on parameter sets in our model where increasing rho decreases $\Rnum$, we may still be ``missing out'' by focusing only on breadth and not depth. }
%% \david{These are thoughtful ideas from Fred.  Perhaps they could be compressed to a couple of sentences describing a future direction worth pursuing.}

@@

% reviewer1:com4 
\rev{Using the expression for the effectiveness of control parameter $\Delta$ in eq.~\ref{eq:del4a}, gives}
\begin{align}
\label{eq:d_del_wis}
\pder \Delta{w\_{IS}}= \dfrac{(\omega-\rho) (\omega(\omega-\rho\theta\_w)+\gamma(\omega-\rho)) (\theta\_w\gamma+\theta\_c\omega p\_I)}
{(-\omega^2 \gamma+\omega \gamma \rho-\gamma \rho \omega w\_{IS}-
\omega \gamma^2+\gamma^2 \rho-\omega^2 p_I \rho w\_{IS})^2},
\end{align}
which is a positive quantity. Thus, $\pder \Rnum{w\_{IS}}\leq 0$. Therefore, increasing the focus of testing on the infectious people will result in less transmission.

>>>>

\newcommand{\fref}[1]{Fig.~\ref{#1}}
\newcommand{\appref}[1]{Appendix~\ref{app:#1}}
% \newcommand{\Rlogo}{R\xspace}
\newcommand{\percap}{\emph{per capita}\xspace}
\newcommand{\Rnum}{\ensuremath{\mathcal{R}_0}\xspace}
\newcommand{\covid}{COVID-19\xspace}
\newcommand{\pro}[1]{\ensuremath{\frac{\partial #1}{\partial \rho}}}
\newcommand{\pder}[2]{\ensuremath{\frac{\partial#1}{\partial#2}}} %\pder y{x}
\newcommand{\testing}[1]{\ensuremath{{\cal T}_{#1}}\xspace}
\newcommand{\testinghat}[1]{\ensuremath{\hat{\cal T}_{#1}}\xspace}
\newcommand{\testtarget}{\ensuremath{w\_{IS}}}
\newcommand{\oldtesttarget}{\ensuremath{\frac{w_I}{w_S}}\xspace}
\newcommand*\subtxt[1]{_{\textnormal{#1}}}
\newcommand{\rev}[1]{\comment{red}{Revised}{#1}}
\renewcommand{\rev}[1]{#1}
%% Roman subscript
\DeclareRobustCommand\_{\ifmmode\expandafter\subtxt\else\textunderscore\fi}

@@

Testing individuals for pathogens can affect the spread of epidemics. Understanding how individual-level processes of sampling and reporting test results can affect community- or population-level spread is a dynamical modeling question.
The effect of testing processes on epidemic dynamics depends on factors underlying implementation, particularly testing intensity and on whom testing is focused.
Here, we use a simple model to explore how the individual-level effects of testing might directly impact population-level spread.
Our model development was motivated by the \covid epidemic, but has generic epidemiological and testing structures. 
To the classic SIR framework we have added a \percap\ testing intensity, and compartment-specific testing weights, which can be adjusted to reflect different testing emphases --- surveillance, diagnosis, or control. 
We derive an analytic expression for the relative reduction in the basic reproductive number due to testing, test-reporting and related isolation behaviours.
Intensive testing and fast test reporting are expected to be beneficial at the community level because they can provide a rapid assessment of the situation, identify hot spots, and may enable rapid contact-tracing. 
Direct effects of fast testing at the individual level are less clear, and may depend on how individuals' behaviour is affected by testing information.
Our simple model shows that under some circumstances both increased testing intensity and faster test reporting can \emph{reduce} the effectiveness of control, and allows us to explore the conditions under which this occurs.
Conversely, we find that focusing testing on infected individuals always acts to increase effectiveness of control.

@@

The Disease-Free Equilibrium (DFE) for the expanded SIR model (Eqs.~\ref{model}) is found by setting the infected compartments to 0 and solving for the unknowns. 
% reviewer1:com3 
The DFE depends on \percap testing intensity $\rho$, the rate of test return $\omega$, and the population size $N$.
The DFE is 
\begin{equation}
\label{dfe}
S\_n^*= \frac{\rho}{\omega} N, \quad S\_u^*= N-S\_n^*, \quad\text{and}\quad I_j=R_j=0 \ \text{for all }j.
\end{equation}
The corresponding \percap testing rate (Eq.~\ref{test}) for the infected compartment $I$ at DFE is one of the key analysis parameters and can be simplified as 
\begin{equation}
\label{eq:fi}
\testinghat{I} = (\omega\rho/(\omega-\rho))w_I/w_S \quad .
\end{equation}

@@

In this paper, we have developed and analyzed a simple compartmental model that combines epidemiological dynamics---as defined by a simple SIR model---with the dynamics of testing and isolation. Our model is a caricature: it models the most basic feedbacks between epidemic and testing processes, but does not attempt to incorporate the many known complications of COVID-19 epidemiology (e.g., exposed, pre-symptomatic, and asymptomatic compartments \citep{kain2021chopping}; time-varying testing rates; behavioural dynamics \citep{weitz2020awareness}); 
% reviewer2:com1
Further, our model ignores the fact that the testing and isolation parameters can change over time due to public health responses. In other words, our model makes an assumption of ``independence'': we assume that testing and isolation affect the epidemic progression, but we do not account for the fact that epidemic progression is also likely to affect testing and isolation strategies.%
Thus, it is most appropriate for assessing the \emph{qualitative} phenomena that arise from the interactions between transmission dynamics and testing, rather than for making quantitative predictions or guiding pandemic responses.

@@

% reviewer1:com1
This finding does, however, suggest that cautious behaviour by people waiting for test results might have appreciable effects on epidemic spread. This reinforces the idea that people waiting for test results should be urged to isolate and take other measures to prevent infecting other, or being infected. It also provides general support for the idea of institutional ``gateway'' testing which restricts individuals' contacts or access until one or more negative tests have been received \citep{muller2021mathematical}.%

@@

Although we model the testing process in more detail than typical epidemiological models, one place where more detail could be informative is in the processes determining the testing weights $\{w_S, w_I, w_R\}$. While random testing, as done for surveillance purposes, unambiguously leads to equal testing weights, making precise quantitative connections between public-health practices and testing weights is difficult in other contexts. The testing weights reflect the correlation between an individual's risk of infection and their likelihood of being tested due to age, occupation, geographic location, etc.. This correlation is influenced, among many other factors, by the proportion of the uninfected population with COVID-like symptoms (e.g., due to seasonal upper respiratory tract infections); the concentration of transmission and testing in hot spots such as long-term care facilities and high-density workplaces; the overall testing intensity (and hence, e.g., restriction to symptomatic individuals); and the proportion of COVID-infected people who are symptomatic.

% reviewer1:com2
Future steps should explore mathematically tractable ways to model some of these factors more precisely. For example, separating the infected class into exposed, symptomatic, and a- or pre-symptomatic compartments and allowing the testing weights to vary across non-symptomatic (exposed/asymptomatic/presymptomatic) vs.\ symptomatic compartments could reflect the allocation of tests for diagnostic purposes (targeting symptomatic individuals) vs.\ contact-tracing (targeting infected but non-symptomatic individuals) vs.\ screening (relatively equal weights, depending on the venue). This could also allow exploration of different assumptions about individuals' likelihood of testing positive in different compartments, with implications for the questions about repeat testing discussed above. Alternatively, one could make the testing weights depend on the testing intensity or test-return rate as suggested above. Whatever complexity is added would probably put the model beyond reach of the analytical methods we have used in this paper, but one could still use semi-numerical methods such as constructing the next-generation matrix and using it to evaluate the derivatives of $\Rnum$ with respect to the parameters numerically.

@@

% reviewer2:com2
%% \fred{ our model makes the counterintuitive suggestion that increased testing rate could increase $\Rnum$. However, note that in our model, we made the assumption that once an individual is in the ``waiting'' compartment, they cannot be tested again until the test result returns. Thus, when the test return time is very long, this creates a massive time-delay where an individual may become infected while waiting for the test result, and then the test result comes back as negative, and so they believe they are negative when they are actually positive (and so they do not self-isolate at all, under the belief that they are negative).
%% So, in this case, it would be beneficial to actually re-test people even if their first test result has not come back. In fact, one could probably conceive of ``testing breadth'' (which roughly corresponds to the number of different people you test over some period of time) vs. ``testing depth (which corresponds to the number of times you test a single person over a period of time). Note that testing breadth gives us granular information ``in space but not in time'' (i.e., it tells us how many people are or are not infected at a single point in time), whereas testing depth gives us granular information ``in time but not in space''. The pathological situation in our model can be viewed as an imbalance between testing depth and breadth, where the testing strategy is overly-broad but not sufficiently deep.
%% In particular, one could build models that differentiate/partition ``testing intensity'' into ``testing breadth'' and ``testing depth''. In such a model, there should always exist some allocation of breadth and depth that makes it so that increasing the overall testing intensity is beneficial (i.e., if we choose the right strategy, more testing should always be good).
%% The most important point of all this: while the exact effect we identified in our model (that more testing can increase $\Rnum$ if done incorrectly) is unlikely to happen in real life, a weaker ``sister effect'' probably happens in real life all the time: we may be allocating our testing resources in the wrong breadth vs. depth ratio. Our model ``proves'' that pure-breadth is not always the optimal strategy (because, in fact, it can increase $\Rnum$). So, a more realistic future question is: what is the optimal allocation between breadth and depth that decreases $\Rnum$ most sharply for a given overall testing intensity? For example, in our model, when we increase testing intensity, we are mostly increasing breadth. So, even on parameter sets in our model where increasing rho decreases $\Rnum$, we may still be ``missing out'' by focusing only on breadth and not depth. }

@@

% reviewer1:com4 
Using the expression for the effectiveness of control parameter $\Delta$ in eq.~\ref{eq:del4a}, gives
\begin{align}
\label{eq:d_del_wis}
\pder \Delta{w\_{IS}}= \dfrac{(\omega-\rho) (\omega(\omega-\rho\theta\_w)+\gamma(\omega-\rho)) (\theta\_w\gamma+\theta\_c\omega p\_I)}
{(-\omega^2 \gamma+\omega \gamma \rho-\gamma \rho \omega w\_{IS}-
\omega \gamma^2+\gamma^2 \rho-\omega^2 p_I \rho w\_{IS})^2},
\end{align}
which is a positive quantity. Thus, $\pder \Rnum{w\_{IS}}\leq 0$. Therefore, increasing the focus of testing on the infectious people will result in less transmission.
